---
title: "Getting started with regulog"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with regulog}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(regulog)
```

## Overview

`regulog` provides tamper-evident, hash-chained audit logging for R
applications. This vignette covers the core API. For Shiny
integration, see `vignette("shiny-integration")`.

## Initialise a session

Every audit log starts with `regulog_init()`. You must supply an application
name and user identity. Optionally provide a `path` to persist entries to disk
as a newline-delimited JSON file (`.rlog`).

```{r init}
log <- regulog_init(
  app     = "clinical-review-tool",
  version = "1.2.0",
  user    = "jsmith"
  # path  = "logs/audit.rlog"  # omit for in-memory only
)

log
```

Without a `path`, the log is in-memory only — useful for development and
testing, but not appropriate for production regulated deployments.

## Logging actions

Use `log_action()` for discrete user actions: approvals, rejections, exports,
sign-offs.

```{r log-action}
log_action(log,
  action = "approved",
  object = "dataset_v3.csv",
  reason = "QC review complete — no anomalies found"
)

log_action(log,
  action = "exported",
  object = "dataset_v3.csv",
  reason = "Sent to statistical analysis team per SAP section 4.1"
)
```

`reason` is **mandatory with no default**. An empty or blank reason raises an
error immediately — every entry in the log must document why it was made.

```{r reason-error, error = TRUE}
# This will error:
log_action(log, action = "approved", object = "file.csv", reason = "")
```

## Logging data changes

Use `log_change()` when a field value is modified. Both the prior and new value
are captured, so you have a complete history of what it was, what it became,
and why.

```{r log-change}
log_change(log,
  object = "patient_10472",
  field  = "dob",
  before = "1985-03-01",
  after  = "1985-03-11",
  reason = "Transcription error — corrected per source document CRF page 14"
)
```

## Verifying chain integrity

`verify_log()` recomputes every entry hash and checks that chain links are
unbroken. Run it at any time, on a schedule, or as part of a handoff.

```{r verify}
verify_log(log)
```

It also works directly from a `.rlog` file path — no R object needed:

```{r verify-file, eval = FALSE}
verify_log("logs/audit.rlog")
```

### What tampering looks like

```{r tamper, error = FALSE, warning = TRUE}
# Simulate tampering — modifying an entry field after the fact
log$entries[[1L]]$action <- "ALTERED"
verify_log(log)
```

## Exporting for regulatory submission

`export_audit_trail()` serialises the log to CSV or JSON. Use `signed = TRUE`
for regulatory submissions — this runs `verify_log()` and stamps `chain_intact`
and `verified_at` into every row.

```{r export-csv}
# Reset to a clean log for the export example
log2 <- regulog_init(app = "clinical-review-tool", version = "1.2.0",
                      user = "jsmith")
log_action(log2, action = "approved", object = "report_final.pdf",
           reason = "All review comments resolved")
log_change(log2, object = "patient_99", field = "visit_date",
           before = "2026-04-01", after = "2026-04-11",
           reason = "Date transposition error per visit log")

df <- export_audit_trail(log2, format = "csv", signed = TRUE)
df
```

### Writing to disk

```{r export-disk, eval = FALSE}
export_audit_trail(log2,
  format = "csv",
  from   = "2026-01-01",
  signed = TRUE,
  path   = "audit_export.csv"
)

# JSON format — useful for archival and system integration
export_audit_trail(log2,
  format = "json",
  signed = TRUE,
  path   = "audit_export.json"
)
```

## The .rlog file format

Entries are written as newline-delimited JSON (one JSON object per line). The
format is deliberately human-readable — a regulator can inspect an audit log
with any text editor:

```
{"entry_id":1,"timestamp":"2026-06-18T14:32:01.123456Z","app":"clinical-review-tool","app_version":"1.2.0","user":"jsmith","type":"ACTION","action":"approved","object":"dataset_v3.csv","reason":"QC review complete","prev_hash":"e3b0c4...","entry_hash":"a87ff6..."}
```

## Validation

`regulog` ships with IQ/OQ/PQ qualification scripts in `inst/validation/`.
Run them in sequence to qualify the package for use in a validated
computerised system:

```{r validation, eval = FALSE}
source(system.file("validation/IQ_regulog.R", package = "regulog"))
source(system.file("validation/OQ_regulog.R", package = "regulog"))
source(system.file("validation/PQ_regulog.R", package = "regulog"))
```

The Requirements Traceability Matrix (RTM) is at:

```{r rtm, eval = FALSE}
rtm_path <- system.file("validation/RTM_regulog.csv", package = "regulog")
read.csv(rtm_path)
```
