---
title: "Audit-ready oncology safety monitoring with regulog"
subtitle: "End-to-end 21 CFR Part 11 audit logging for a Phase III NSCLC trial"
author: "Ndoh Penn"
date: "2026-06-26"
categories: [regulog, oncology, safety, 21-CFR-Part-11, audit-trail, TEAE]
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-tools: true
    highlight-style: github
execute:
  echo: true
  warning: false
  message: false
---

```{r setup, include=FALSE}
set.seed(2026)
knitr::opts_chunk$set(comment = "#>", fig.width = 7, fig.height = 4)
```

## Overview

Oncology safety monitoring is one of the highest-scrutiny activities in clinical development. Every data cut decision, every SAE narrative, every judgment about treatment-relatedness, and every action taken in response to a safety signal must be traceable, documented, and available for inspection.

In practice, safety monitoring produces a trail of emails, meeting minutes, and retrospective notes that is difficult to assemble and impossible to cryptographically verify. `regulog` replaces this with a hash-chained audit trail that is built as the analysis runs — not reconstructed after the fact.

This article walks through a complete safety monitoring workflow for a simulated Phase III non-small cell lung cancer (NSCLC) trial, covering TEAE summaries, Grade 3+ adverse event analysis, SAE flagging, safety signal documentation, medical monitor sign-off, and submission-ready audit trail export.

---

## Setup

```{r packages}
library(regulog)
library(dplyr)
library(tidyr)
library(ggplot2)
```

---

## Simulate the safety datasets

We simulate CDISC ADaM-structured ADSL and ADAE datasets for a Phase III trial of a novel KRAS G12C inhibitor versus docetaxel in second-line NSCLC. 400 randomised patients, 200 per arm.

```{r simulate}
n_per_arm <- 200L
n_total   <- n_per_arm * 2L

# ── ADSL ──────────────────────────────────────────────────────────────────
adsl <- data.frame(
  STUDYID = "NSCLC-KRASi-301",
  USUBJID = sprintf("KRASi301-%04d", seq_len(n_total)),
  TRT01P  = rep(c("KRASi 400mg QD", "Docetaxel 75mg/m2"), each = n_per_arm),
  TRT01PN = rep(c(1L, 2L), each = n_per_arm),
  AGE     = round(rnorm(n_total, mean = 63, sd = 9)),
  SEX     = sample(c("M", "F"), n_total, replace = TRUE, prob = c(0.52, 0.48)),
  ITTFL   = "Y",
  SAFFL   = "Y",
  EOSSTT  = sample(c("COMPLETED", "DISCONTINUED", "ONGOING"),
                   n_total, replace = TRUE, prob = c(0.35, 0.45, 0.20)),
  stringsAsFactors = FALSE
)

# ── Adverse event system organ classes ────────────────────────────────────
socs <- c(
  "Gastrointestinal disorders",
  "General disorders and administration site conditions",
  "Skin and subcutaneous tissue disorders",
  "Respiratory, thoracic and mediastinal disorders",
  "Metabolism and nutrition disorders",
  "Blood and lymphatic system disorders",
  "Nervous system disorders",
  "Infections and infestations"
)

ae_terms <- list(
  "Gastrointestinal disorders"                          = c("Nausea", "Diarrhoea", "Vomiting", "Constipation"),
  "General disorders and administration site conditions" = c("Fatigue", "Pyrexia", "Oedema peripheral"),
  "Skin and subcutaneous tissue disorders"              = c("Rash", "Pruritus", "Dry skin"),
  "Respiratory, thoracic and mediastinal disorders"     = c("Cough", "Dyspnoea", "Pneumonitis"),
  "Metabolism and nutrition disorders"                  = c("Decreased appetite", "Hyponatraemia", "Hyperglycaemia"),
  "Blood and lymphatic system disorders"                = c("Anaemia", "Neutropenia", "Thrombocytopenia"),
  "Nervous system disorders"                            = c("Headache", "Peripheral neuropathy", "Dizziness"),
  "Infections and infestations"                         = c("Upper respiratory tract infection", "Pneumonia")
)

# Simulate AEs — KRASi has more GI/skin; docetaxel has more haematologic
set.seed(2026)
adae_list <- lapply(seq_len(n_total), function(i) {
  subj <- adsl$USUBJID[i]
  trt  <- adsl$TRT01P[i]
  is_krasi <- trt == "KRASi 400mg QD"

  n_ae <- rpois(1, lambda = if (is_krasi) 4.2 else 3.8)
  if (n_ae == 0L) return(NULL)

  # Weighted SOC sampling
  soc_weights <- if (is_krasi) {
    c(0.30, 0.18, 0.20, 0.08, 0.08, 0.04, 0.06, 0.06)
  } else {
    c(0.18, 0.16, 0.06, 0.10, 0.10, 0.22, 0.08, 0.10)
  }

  sampled_socs <- sample(socs, n_ae, replace = TRUE, prob = soc_weights)

  aes <- do.call(rbind, lapply(seq_along(sampled_socs), function(j) {
    soc  <- sampled_socs[j]
    term <- sample(ae_terms[[soc]], 1L)

    # Grade: KRASi more Grade 2; docetaxel more Grade 3/4 haematologic
    grade_probs <- if (is_krasi && soc == "Blood and lymphatic system disorders") {
      c(0.4, 0.35, 0.20, 0.05)
    } else if (!is_krasi && soc == "Blood and lymphatic system disorders") {
      c(0.2, 0.30, 0.35, 0.15)
    } else {
      c(0.45, 0.35, 0.16, 0.04)
    }
    grade <- sample(1:4, 1L, prob = grade_probs)

    data.frame(
      STUDYID  = "NSCLC-KRASi-301",
      USUBJID  = subj,
      TRT01P   = trt,
      AEBODSYS = soc,
      AEDECOD  = term,
      AESEV    = c("MILD", "MODERATE", "SEVERE", "LIFE-THREATENING")[grade],
      AETOXGR  = as.character(grade),
      AESER    = ifelse(grade >= 3 & runif(1) < 0.35, "Y", "N"),
      AEREL    = sample(c("RELATED", "NOT RELATED", "POSSIBLY RELATED"),
                        1L, prob = c(0.5, 0.25, 0.25)),
      AEACN    = ifelse(grade >= 3,
                        sample(c("DOSE REDUCED", "DRUG WITHDRAWN", "DOSE NOT CHANGED"),
                               1L, prob = c(0.4, 0.3, 0.3)),
                        "DOSE NOT CHANGED"),
      AESTDY   = sample(1:365, 1L),
      stringsAsFactors = FALSE
    )
  }))
  aes
})

adae <- do.call(rbind, Filter(Negate(is.null), adae_list))
adae <- adae[order(adae$USUBJID, adae$AESTDY), ]
rownames(adae) <- NULL

cat(sprintf(
  "ADSL: %d subjects | ADAE: %d AEs across %d subjects\n",
  nrow(adsl),
  nrow(adae),
  n_distinct(adae$USUBJID)
))
```

---

## Open the safety monitoring session

```{r session}
lg <- rl_session(
  analyst  = "Ndoh Penn",
  analysis = "NSCLC-KRASi-301 Integrated Safety Summary — Data Cut 2026-05-15",
  reason   = "Scheduled safety data cut for Safety Monitoring Committee (SMC) review",
  system   = list(
    protocol     = "NSCLC-KRASi-301",
    data_cut     = "2026-05-15",
    analysis_set = "Safety analysis set (SAFFL = Y)",
    smc_meeting  = "2026-07-10",
    regulatory   = "21 CFR Part 11, ICH E6(R2)"
  )
)

lg
```

---

## Log data access and data cut

```{r data-load}
rl_action(lg,
  action = "LOAD_ADSL",
  reason = "Load subject-level dataset for safety population definition",
  detail = sprintf("ADSL: %d subjects", nrow(adsl))
)

rl_action(lg,
  action = "LOAD_ADAE",
  reason = "Load adverse event dataset for safety analysis",
  detail = sprintf("ADAE: %d records from %d subjects",
                   nrow(adae), n_distinct(adae$USUBJID))
)

# Document the data cut — this is a critical audit point
rl_note(lg,
  action = "DATA_CUT_CONFIRMED",
  detail = "Data cut date confirmed as 2026-05-15 per DMC charter Section 4.2.
            All AEs with onset on or before 2026-05-15 included.
            Lock confirmed by DM team (ref: DM-lock-20260518-001)"
)
```

---

## Apply safety analysis population

```{r population}
# Safety analysis set
adsl_saf <- adsl |> filter(SAFFL == "Y")
adae_saf <- adae |> filter(USUBJID %in% adsl_saf$USUBJID)

arm_n <- adsl_saf |> count(TRT01P)

rl_action(lg,
  action = "DEFINE_SAFETY_SET",
  reason = "Apply safety analysis set (SAFFL = Y) per SAP Section 4.1",
  detail = paste(sprintf("%s: n = %d", arm_n$TRT01P, arm_n$n), collapse = " | ")
)
```

---

## TEAE incidence summary

```{r teae-summary}
# Subjects with at least one TEAE by SOC and treatment
teae_summary <- adae_saf |>
  group_by(TRT01P, AEBODSYS) |>
  summarise(
    n_subjects = n_distinct(USUBJID),
    n_events   = n(),
    .groups    = "drop"
  ) |>
  left_join(arm_n, by = "TRT01P") |>
  mutate(pct = round(n_subjects / n * 100, 1)) |>
  arrange(AEBODSYS, TRT01P)

rl_action(lg,
  action = "COMPUTE_TEAE_INCIDENCE",
  reason = "Compute TEAE incidence by SOC and treatment arm per SAP Section 6.1",
  detail = sprintf(
    "TEAEs reported: %d subjects in KRASi arm (%.1f%%), %d in docetaxel arm (%.1f%%)",
    teae_summary$n_subjects[teae_summary$TRT01P == "KRASi 400mg QD"] |> max(),
    teae_summary$pct[teae_summary$TRT01P == "KRASi 400mg QD"] |> max(),
    teae_summary$n_subjects[teae_summary$TRT01P == "Docetaxel 75mg/m2"] |> max(),
    teae_summary$pct[teae_summary$TRT01P == "Docetaxel 75mg/m2"] |> max()
  )
)

# Top 5 SOCs by overall incidence
top_socs <- teae_summary |>
  group_by(AEBODSYS) |>
  summarise(total = sum(n_subjects)) |>
  slice_max(total, n = 5L) |>
  pull(AEBODSYS)

teae_summary |>
  filter(AEBODSYS %in% top_socs) |>
  select(SOC = AEBODSYS, Treatment = TRT01P,
         `N subjects` = n_subjects, `%` = pct) |>
  knitr::kable(caption = "TEAE incidence by SOC — top 5 (safety analysis set)")
```

---

## Grade 3/4 adverse events

```{r grade34}
g34 <- adae_saf |>
  filter(AETOXGR %in% c("3", "4")) |>
  group_by(TRT01P, AEBODSYS, AEDECOD, AETOXGR) |>
  summarise(n = n_distinct(USUBJID), .groups = "drop") |>
  left_join(arm_n, by = "TRT01P") |>
  mutate(pct = round(n / n.y * 100, 1)) |>
  arrange(desc(n))

rl_action(lg,
  action = "COMPUTE_GRADE34_AE",
  reason = "Compute Grade 3/4 TEAE incidence per SAP Section 6.2 and ICH E9",
  detail = sprintf(
    "Grade 3/4 AEs: %d events in KRASi arm, %d in docetaxel arm",
    nrow(adae_saf[adae_saf$TRT01P == "KRASi 400mg QD" &
                  adae_saf$AETOXGR %in% c("3","4"), ]),
    nrow(adae_saf[adae_saf$TRT01P == "Docetaxel 75mg/m2" &
                  adae_saf$AETOXGR %in% c("3","4"), ])
  )
)

g34 |>
  select(Treatment = TRT01P, SOC = AEBODSYS,
         Term = AEDECOD, Grade = AETOXGR,
         `N subjects` = n, `%` = pct) |>
  head(12L) |>
  knitr::kable(caption = "Grade 3/4 TEAEs — top 12 (safety analysis set)")
```

---

## Serious adverse events

```{r saes}
saes <- adae_saf |>
  filter(AESER == "Y") |>
  group_by(TRT01P, AEBODSYS, AEDECOD) |>
  summarise(n = n_distinct(USUBJID), .groups = "drop") |>
  left_join(arm_n, by = "TRT01P") |>
  mutate(pct = round(n / n * 100, 1)) |>
  arrange(desc(n))

n_sae_krasi <- n_distinct(
  adae_saf$USUBJID[adae_saf$AESER == "Y" & adae_saf$TRT01P == "KRASi 400mg QD"]
)
n_sae_doce  <- n_distinct(
  adae_saf$USUBJID[adae_saf$AESER == "Y" & adae_saf$TRT01P == "Docetaxel 75mg/m2"]
)

rl_action(lg,
  action = "COMPUTE_SAE_SUMMARY",
  reason = "Summarise serious adverse events per SAP Section 6.3",
  detail = sprintf(
    "SAEs: %d subjects in KRASi arm (%.1f%%), %d in docetaxel arm (%.1f%%)",
    n_sae_krasi,
    n_sae_krasi / n_per_arm * 100,
    n_sae_doce,
    n_sae_doce / n_per_arm * 100
  )
)

saes |>
  select(Treatment = TRT01P, SOC = AEBODSYS, Term = AEDECOD, `N subjects` = n) |>
  head(10L) |>
  knitr::kable(caption = "Serious TEAEs (AESER = Y) — top 10")
```

---

## Pneumonitis signal review

Pneumonitis is a clinically important immune-mediated AE associated with KRAS inhibitors. We identify cases and document the medical review decision.

```{r pneumonitis}
# Identify pneumonitis cases
pneumonitis <- adae_saf |>
  filter(grepl("Pneumonitis", AEDECOD, ignore.case = TRUE))

n_pneu_krasi <- nrow(pneumonitis[pneumonitis$TRT01P == "KRASi 400mg QD", ])
n_pneu_doce  <- nrow(pneumonitis[pneumonitis$TRT01P == "Docetaxel 75mg/m2", ])
n_pneu_g3    <- nrow(pneumonitis[pneumonitis$AETOXGR %in% c("3","4"), ])

rl_note(lg,
  action = "SAFETY_SIGNAL_IDENTIFIED",
  detail = sprintf(
    "Pneumonitis signal identified: %d cases in KRASi arm vs %d in docetaxel arm.
     Grade 3/4: %d cases. Flagged for medical monitor review per Safety Review Plan.",
    n_pneu_krasi, n_pneu_doce, n_pneu_g3
  )
)

# Log individual Grade 3+ pneumonitis cases
g3_pneu <- pneumonitis |> filter(AETOXGR %in% c("3","4"))
for (i in seq_len(nrow(g3_pneu))) {
  rl_note(lg,
    action = "SAE_PNEUMONITIS_CASE",
    detail = sprintf(
      "USUBJID %s | Grade %s | Treatment: %s | Action: %s | Related: %s",
      g3_pneu$USUBJID[i], g3_pneu$AETOXGR[i],
      g3_pneu$TRT01P[i], g3_pneu$AEACN[i], g3_pneu$AEREL[i]
    )
  )
}

# Log medical monitor decision
rl_note(lg,
  action  = "MEDICAL_MONITOR_DECISION",
  detail  = sprintf(
    "Medical monitor reviewed all %d pneumonitis cases (ref: MM-review-20260620-001).
     Conclusion: Incidence consistent with class effect; benefit-risk remains favourable.
     Risk mitigation: enhanced monitoring protocol added to IB (v4.2, dated 2026-06-22).
     No protocol amendment required at this time.",
    n_pneu_krasi
  )
)

pneumonitis |>
  select(USUBJID, Treatment = TRT01P, Term = AEDECOD,
         Grade = AETOXGR, Serious = AESER,
         Related = AEREL, Action = AEACN) |>
  knitr::kable(caption = "Pneumonitis cases — all grades")
```

---

## Discontinuations due to AEs

```{r discontinuations}
disc_ae <- adae_saf |>
  filter(AEACN == "DRUG WITHDRAWN") |>
  group_by(TRT01P, AEBODSYS, AEDECOD) |>
  summarise(n = n_distinct(USUBJID), .groups = "drop") |>
  arrange(desc(n))

n_disc_krasi <- n_distinct(
  adae_saf$USUBJID[adae_saf$AEACN == "DRUG WITHDRAWN" &
                   adae_saf$TRT01P == "KRASi 400mg QD"]
)
n_disc_doce <- n_distinct(
  adae_saf$USUBJID[adae_saf$AEACN == "DRUG WITHDRAWN" &
                   adae_saf$TRT01P == "Docetaxel 75mg/m2"]
)

rl_action(lg,
  action = "COMPUTE_DISC_DUE_TO_AE",
  reason = "Tabulate treatment discontinuations due to AE per SAP Section 6.4",
  detail = sprintf(
    "Discontinued due to AE: %d subjects in KRASi arm (%.1f%%), %d in docetaxel (%.1f%%)",
    n_disc_krasi, n_disc_krasi / n_per_arm * 100,
    n_disc_doce,  n_disc_doce  / n_per_arm * 100
  )
)

disc_ae |>
  select(Treatment = TRT01P, SOC = AEBODSYS,
         Term = AEDECOD, `N subjects` = n) |>
  head(10L) |>
  knitr::kable(caption = "Discontinuations due to AE — top 10")
```

---

## Safety visualisations

```{r forest-plot}
# Treatment-emergent AE incidence by SOC
teae_wide <- teae_summary |>
  filter(AEBODSYS %in% top_socs) |>
  select(TRT01P, AEBODSYS, pct) |>
  pivot_wider(names_from = TRT01P, values_from = pct, values_fill = 0) |>
  rename(KRASi = `KRASi 400mg QD`, Docetaxel = `Docetaxel 75mg/m2`)

teae_wide |>
  tidyr::pivot_longer(cols = c(KRASi, Docetaxel),
                      names_to = "Treatment", values_to = "pct") |>
  ggplot(aes(x = pct, y = reorder(AEBODSYS, pct), fill = Treatment)) +
  geom_col(position = "dodge", width = 0.65, alpha = 0.85) +
  scale_fill_manual(values = c("KRASi" = "#1a56db", "Docetaxel" = "#6b6f80")) +
  labs(
    title    = "TEAE Incidence by System Organ Class",
    subtitle = "Safety analysis set | Top 5 SOCs",
    x        = "Subjects with ≥1 TEAE (%)",
    y        = NULL,
    fill     = "Treatment"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")

rl_note(lg,
  action = "FIGURE_GENERATED",
  detail = "Figure 1: TEAE incidence by SOC (top 5), dodge bar chart"
)
```

```{r grade-plot}
# Grade distribution by treatment arm
adae_saf |>
  count(TRT01P, AETOXGR) |>
  group_by(TRT01P) |>
  mutate(pct = n / sum(n) * 100) |>
  ggplot(aes(x = AETOXGR, y = pct, fill = TRT01P)) +
  geom_col(position = "dodge", width = 0.65, alpha = 0.85) +
  scale_fill_manual(values = c("KRASi 400mg QD" = "#1a56db",
                               "Docetaxel 75mg/m2" = "#6b6f80")) +
  labs(
    title    = "AE Grade Distribution by Treatment Arm",
    subtitle = "Safety analysis set | % of all AEs",
    x        = "Grade (CTCAE)",
    y        = "% of AEs",
    fill     = "Treatment"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")

rl_note(lg,
  action = "FIGURE_GENERATED",
  detail = "Figure 2: AE grade distribution by treatment arm"
)
```

---

## Safety summary for SMC

```{r smc-summary}
smc_table <- data.frame(
  Endpoint = c(
    "Subjects with ≥1 TEAE",
    "Subjects with ≥1 Grade 3/4 TEAE",
    "Subjects with ≥1 SAE",
    "Discontinuation due to AE",
    "Pneumonitis (all grades)",
    "Pneumonitis (Grade 3/4)"
  ),
  KRASi = c(
    sprintf("%d (%.1f%%)", n_distinct(adae_saf$USUBJID[adae_saf$TRT01P=="KRASi 400mg QD"]),
            n_distinct(adae_saf$USUBJID[adae_saf$TRT01P=="KRASi 400mg QD"]) / n_per_arm * 100),
    sprintf("%d (%.1f%%)", n_distinct(adae_saf$USUBJID[adae_saf$TRT01P=="KRASi 400mg QD" & adae_saf$AETOXGR %in% c("3","4")]),
            n_distinct(adae_saf$USUBJID[adae_saf$TRT01P=="KRASi 400mg QD" & adae_saf$AETOXGR %in% c("3","4")]) / n_per_arm * 100),
    sprintf("%d (%.1f%%)", n_sae_krasi,  n_sae_krasi  / n_per_arm * 100),
    sprintf("%d (%.1f%%)", n_disc_krasi, n_disc_krasi / n_per_arm * 100),
    sprintf("%d (%.1f%%)", n_pneu_krasi, n_pneu_krasi / n_per_arm * 100),
    sprintf("%d (%.1f%%)", n_pneu_g3,    n_pneu_g3    / n_per_arm * 100)
  ),
  Docetaxel = c(
    sprintf("%d (%.1f%%)", n_distinct(adae_saf$USUBJID[adae_saf$TRT01P=="Docetaxel 75mg/m2"]),
            n_distinct(adae_saf$USUBJID[adae_saf$TRT01P=="Docetaxel 75mg/m2"]) / n_per_arm * 100),
    sprintf("%d (%.1f%%)", n_distinct(adae_saf$USUBJID[adae_saf$TRT01P=="Docetaxel 75mg/m2" & adae_saf$AETOXGR %in% c("3","4")]),
            n_distinct(adae_saf$USUBJID[adae_saf$TRT01P=="Docetaxel 75mg/m2" & adae_saf$AETOXGR %in% c("3","4")]) / n_per_arm * 100),
    sprintf("%d (%.1f%%)", n_sae_doce,  n_sae_doce  / n_per_arm * 100),
    sprintf("%d (%.1f%%)", n_disc_doce, n_disc_doce / n_per_arm * 100),
    sprintf("%d (%.1f%%)", n_pneu_doce, n_pneu_doce / n_per_arm * 100),
    "0 (0.0%)"
  ),
  stringsAsFactors = FALSE
)

knitr::kable(smc_table,
  col.names = c("Safety endpoint",
                "KRASi 400mg QD (N=200)",
                "Docetaxel 75mg/m2 (N=200)"),
  caption   = "Safety summary — NSCLC-KRASi-301 (data cut 2026-05-15)"
)

rl_action(lg,
  action = "SMC_SAFETY_TABLE_COMPILED",
  reason = "Compile integrated safety summary for Safety Monitoring Committee",
  detail = "All 6 pre-specified safety endpoints computed and tabulated per SMC charter"
)
```

---

## Dual electronic sign-off

Oncology safety reports require both biostatistician and medical monitor signatures. `regulog` supports sequential sign-offs in a single hash chain:

```{r signatures}
# Biostatistician sign-off
rl_signature(lg,
  signatory = "Ndoh Penn",
  role      = "Lead Biostatistician",
  meaning   = "I confirm that the safety analyses were conducted in accordance
               with SAP v2.1 and the SMC charter. All AEs, SAEs, and
               discontinuations are correctly classified and tabulated.",
  date      = Sys.Date()
)

# Medical monitor sign-off
rl_signature(lg,
  signatory = "Medical Monitor",
  role      = "Medical Monitor",
  meaning   = "I confirm that I have reviewed the safety data including all
               pneumonitis cases and the integrated safety summary. The
               benefit-risk profile remains favourable for KRASi 400mg QD.",
  date      = Sys.Date()
)
```

---

## Verify and export

```{r verify-export}
# Verify the complete hash chain
rl_verify(lg)

# Export for submission package
trail <- rl_export(lg,
  path   = file.path(tempdir(), "audit_KRASi301_safety_v1.csv"),
  format = "csv"
)

# Preview
head(as.data.frame(lg), 12L) |>
  select(timestamp, action, detail) |>
  knitr::kable(caption = "Audit trail — first 12 entries")
```

---

## What this delivers to regulators

The exported audit trail documents the complete safety monitoring process:

- **Data provenance** — what datasets were loaded, from where, at what data cut
- **Population decisions** — who was included in the safety set and why
- **Analysis execution** — every TEAE table, Grade 3/4 summary, and SAE listing was logged at computation
- **Safety signal handling** — the pneumonitis signal was identified, individually documented, reviewed by the medical monitor, and a decision reached — all in the same hash chain
- **Dual sign-off** — both the biostatistician and medical monitor signatures are cryptographically linked to the analysis that produced the results
- **Tamper evidence** — any post-hoc modification to any entry breaks the SHA-256 chain, detectable without reference to the original system

This directly satisfies 21 CFR Part 11 §11.10(b) (audit trails), §11.10(e) (tamper evidence), §11.100 (electronic signature components), and EU Annex 11 Clause 9 (audit trail) — not through a parallel documentation process, but as an inherent property of the analysis itself.

---

## Session information

```{r session-info}
sessionInfo()
```
