---
title: "Audit-ready oncology safety monitoring with regulog"
author: "Ndoh Penn"
date: "2026-06-26"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Audit-ready oncology safety monitoring with regulog}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include=FALSE}
set.seed(2026)
knitr::opts_chunk$set(comment = "#>", fig.width = 7, fig.height = 4)
```

## Overview

Oncology safety monitoring is one of the highest-scrutiny activities in clinical development. Every data cut decision, every SAE narrative, every judgment about treatment-relatedness, and every action taken in response to a safety signal must be traceable, documented, and available for inspection.

In practice, safety monitoring produces a trail of emails, meeting minutes, and retrospective notes that is difficult to assemble and impossible to cryptographically verify. `regulog` replaces this with a hash-chained audit trail that is built as the analysis runs — not reconstructed after the fact.

This article walks through a complete safety monitoring workflow for a simulated Phase III non-small cell lung cancer (NSCLC) trial, covering TEAE summaries, Grade 3+ adverse event analysis, SAE flagging, safety signal documentation, dual medical/statistical sign-off, and submission-ready audit trail export.

---

## Setup

```{r packages}
library(regulog)
library(dplyr)
library(tidyr)
library(ggplot2)
```

---

## Simulate the safety datasets

We simulate CDISC ADaM-structured ADSL and ADAE datasets for a Phase III trial of a novel KRAS G12C inhibitor versus docetaxel in second-line NSCLC. 400 randomised patients, 200 per arm.

```{r simulate}
n_per_arm <- 200L
n_total   <- n_per_arm * 2L

# ── ADSL ──────────────────────────────────────────────────────────────────
adsl <- data.frame(
  STUDYID = "NSCLC-KRASi-301",
  USUBJID = sprintf("KRASi301-%04d", seq_len(n_total)),
  TRT01P  = rep(c("KRASi 400mg QD", "Docetaxel 75mg/m2"), each = n_per_arm),
  TRT01PN = rep(c(1L, 2L), each = n_per_arm),
  AGE     = round(rnorm(n_total, mean = 63, sd = 9)),
  SEX     = sample(c("M", "F"), n_total, replace = TRUE, prob = c(0.52, 0.48)),
  ITTFL   = "Y",
  SAFFL   = "Y",
  EOSSTT  = sample(c("COMPLETED", "DISCONTINUED", "ONGOING"),
                   n_total, replace = TRUE, prob = c(0.35, 0.45, 0.20)),
  stringsAsFactors = FALSE
)

# ── Adverse event system organ classes ────────────────────────────────────
socs <- c(
  "Gastrointestinal disorders",
  "General disorders and administration site conditions",
  "Skin and subcutaneous tissue disorders",
  "Respiratory, thoracic and mediastinal disorders",
  "Metabolism and nutrition disorders",
  "Blood and lymphatic system disorders",
  "Nervous system disorders",
  "Infections and infestations"
)

ae_terms <- list(
  "Gastrointestinal disorders"                          = c("Nausea", "Diarrhoea", "Vomiting", "Constipation"),
  "General disorders and administration site conditions" = c("Fatigue", "Pyrexia", "Oedema peripheral"),
  "Skin and subcutaneous tissue disorders"              = c("Rash", "Pruritus", "Dry skin"),
  "Respiratory, thoracic and mediastinal disorders"     = c("Cough", "Dyspnoea", "Pneumonitis"),
  "Metabolism and nutrition disorders"                  = c("Decreased appetite", "Hyponatraemia", "Hyperglycaemia"),
  "Blood and lymphatic system disorders"                = c("Anaemia", "Neutropenia", "Thrombocytopenia"),
  "Nervous system disorders"                            = c("Headache", "Peripheral neuropathy", "Dizziness"),
  "Infections and infestations"                         = c("Upper respiratory tract infection", "Pneumonia")
)

# Simulate AEs — KRASi has more GI/skin; docetaxel has more haematologic
set.seed(2026)
adae_list <- lapply(seq_len(n_total), function(i) {
  subj <- adsl$USUBJID[i]
  trt  <- adsl$TRT01P[i]
  is_krasi <- trt == "KRASi 400mg QD"

  n_ae <- rpois(1, lambda = if (is_krasi) 4.2 else 3.8)
  if (n_ae == 0L) return(NULL)

  soc_weights <- if (is_krasi) {
    c(0.30, 0.18, 0.20, 0.08, 0.08, 0.04, 0.06, 0.06)
  } else {
    c(0.18, 0.16, 0.06, 0.10, 0.10, 0.22, 0.08, 0.10)
  }

  sampled_socs <- sample(socs, n_ae, replace = TRUE, prob = soc_weights)

  aes <- do.call(rbind, lapply(seq_along(sampled_socs), function(j) {
    soc  <- sampled_socs[j]
    term <- sample(ae_terms[[soc]], 1L)

    grade_probs <- if (is_krasi && soc == "Blood and lymphatic system disorders") {
      c(0.4, 0.35, 0.20, 0.05)
    } else if (!is_krasi && soc == "Blood and lymphatic system disorders") {
      c(0.2, 0.30, 0.35, 0.15)
    } else {
      c(0.45, 0.35, 0.16, 0.04)
    }
    grade <- sample(1:4, 1L, prob = grade_probs)

    data.frame(
      STUDYID  = "NSCLC-KRASi-301",
      USUBJID  = subj,
      TRT01P   = trt,
      AEBODSYS = soc,
      AEDECOD  = term,
      AESEV    = c("MILD", "MODERATE", "SEVERE", "LIFE-THREATENING")[grade],
      AETOXGR  = as.character(grade),
      AESER    = ifelse(grade >= 3 & runif(1) < 0.35, "Y", "N"),
      AEREL    = sample(c("RELATED", "NOT RELATED", "POSSIBLY RELATED"),
                        1L, prob = c(0.5, 0.25, 0.25)),
      AEACN    = ifelse(grade >= 3,
                        sample(c("DOSE REDUCED", "DRUG WITHDRAWN", "DOSE NOT CHANGED"),
                               1L, prob = c(0.4, 0.3, 0.3)),
                        "DOSE NOT CHANGED"),
      AESTDY   = sample(1:365, 1L),
      stringsAsFactors = FALSE
    )
  }))
  aes
})

adae <- do.call(rbind, Filter(Negate(is.null), adae_list))
adae <- adae[order(adae$USUBJID, adae$AESTDY), ]
rownames(adae) <- NULL

cat(sprintf(
  "ADSL: %d subjects | ADAE: %d AEs across %d subjects\n",
  nrow(adsl),
  nrow(adae),
  n_distinct(adae$USUBJID)
))

# Write to disk so rl_read()/with_log() have real files to read
tmp_adsl <- tempfile(fileext = ".csv")
tmp_adae <- tempfile(fileext = ".csv")
write.csv(adsl, tmp_adsl, row.names = FALSE)
write.csv(adae, tmp_adae, row.names = FALSE)
```

---

## Open the safety monitoring session

`regulog_init()` creates the session and writes the genesis record immediately. The data cut and protocol context are captured as the first logged note, since they are study metadata rather than session configuration.

```{r session}
log <- regulog_init(
  app     = "NSCLC-KRASi-301-safety-summary",
  version = "1.0.0",
  user    = "jsmith",
  path    = file.path(tempdir(), "audit_KRASi301_safety_v1.rlog")
)

log_note(log,
  "Scheduled safety data cut for Safety Monitoring Committee (SMC) review.
   Protocol: NSCLC-KRASi-301. Data cut: 2026-05-15. Analysis set: safety
   analysis set (SAFFL = Y). SMC meeting: 2026-07-10. Regulatory basis:
   21 CFR Part 11, ICH E6(R2)."
)

log
```

---

## Log data access and the data cut

Loading ADSL and ADAE is logged automatically with `with_log()` — row and column counts are captured at the moment of read, not reconstructed afterwards.

```{r data-load}
with_log(log, {
  adsl_loaded <- read(read.csv, tmp_adsl)
  adae_loaded <- read(read.csv, tmp_adae)
})

# Document the data cut — a critical audit point in any safety review
log_note(log,
  "Data cut date confirmed as 2026-05-15 per DMC charter Section 4.2.
   All AEs with onset on or before 2026-05-15 included.
   Lock confirmed by DM team (ref: DM-lock-20260518-001)."
)
```

---

## Apply safety analysis population

```{r population}
adsl_saf <- adsl |> filter(SAFFL == "Y")
adae_saf <- adae |> filter(USUBJID %in% adsl_saf$USUBJID)

arm_n <- adsl_saf |> count(TRT01P)

log_action(log,
  action = "define_safety_set",
  object = "NSCLC-KRASi-301 safety analysis set",
  reason = sprintf(
    "Applied safety analysis set (SAFFL = Y) per SAP Section 4.1. %s",
    paste(sprintf("%s: n = %d", arm_n$TRT01P, arm_n$n), collapse = " | ")
  )
)
```

---

## TEAE incidence summary

```{r teae-summary}
teae_summary <- adae_saf |>
  group_by(TRT01P, AEBODSYS) |>
  summarise(
    n_subjects = n_distinct(USUBJID),
    n_events   = n(),
    .groups    = "drop"
  ) |>
  left_join(arm_n, by = "TRT01P") |>
  mutate(pct = round(n_subjects / n * 100, 1)) |>
  arrange(AEBODSYS, TRT01P)

log_action(log,
  action = "compute_teae_incidence",
  object = "TEAE incidence table by SOC",
  reason = sprintf(
    "Computed per SAP Section 6.1. TEAEs reported: %d subjects in KRASi arm (%.1f%%), %d in docetaxel arm (%.1f%%)",
    teae_summary$n_subjects[teae_summary$TRT01P == "KRASi 400mg QD"] |> max(),
    teae_summary$pct[teae_summary$TRT01P == "KRASi 400mg QD"] |> max(),
    teae_summary$n_subjects[teae_summary$TRT01P == "Docetaxel 75mg/m2"] |> max(),
    teae_summary$pct[teae_summary$TRT01P == "Docetaxel 75mg/m2"] |> max()
  )
)

top_socs <- teae_summary |>
  group_by(AEBODSYS) |>
  summarise(total = sum(n_subjects)) |>
  slice_max(total, n = 5L) |>
  pull(AEBODSYS)

teae_summary |>
  filter(AEBODSYS %in% top_socs) |>
  select(SOC = AEBODSYS, Treatment = TRT01P,
         `N subjects` = n_subjects, `%` = pct) |>
  knitr::kable(caption = "TEAE incidence by SOC — top 5 (safety analysis set)")
```

---

## Grade 3/4 adverse events

```{r grade34}
g34 <- adae_saf |>
  filter(AETOXGR %in% c("3", "4")) |>
  group_by(TRT01P, AEBODSYS, AEDECOD, AETOXGR) |>
  summarise(n = n_distinct(USUBJID), .groups = "drop") |>
  left_join(arm_n |> rename(arm_total = n), by = "TRT01P") |>
  mutate(pct = round(n / arm_total * 100, 1)) |>
  arrange(desc(n))

log_action(log,
  action = "compute_grade34_ae",
  object = "Grade 3/4 TEAE table",
  reason = sprintf(
    "Computed per SAP Section 6.2 and ICH E9. Grade 3/4 AEs: %d events in KRASi arm, %d in docetaxel arm",
    nrow(adae_saf[adae_saf$TRT01P == "KRASi 400mg QD" &
                  adae_saf$AETOXGR %in% c("3","4"), ]),
    nrow(adae_saf[adae_saf$TRT01P == "Docetaxel 75mg/m2" &
                  adae_saf$AETOXGR %in% c("3","4"), ])
  )
)

g34 |>
  select(Treatment = TRT01P, SOC = AEBODSYS,
         Term = AEDECOD, Grade = AETOXGR,
         `N subjects` = n, `%` = pct) |>
  head(12L) |>
  knitr::kable(caption = "Grade 3/4 TEAEs — top 12 (safety analysis set)")
```

---

## Serious adverse events

```{r saes}
saes <- adae_saf |>
  filter(AESER == "Y") |>
  group_by(TRT01P, AEBODSYS, AEDECOD) |>
  summarise(n = n_distinct(USUBJID), .groups = "drop") |>
  left_join(arm_n |> rename(arm_total = n), by = "TRT01P") |>
  mutate(pct = round(n / arm_total * 100, 1)) |>
  arrange(desc(n))

n_sae_krasi <- n_distinct(
  adae_saf$USUBJID[adae_saf$AESER == "Y" & adae_saf$TRT01P == "KRASi 400mg QD"]
)
n_sae_doce  <- n_distinct(
  adae_saf$USUBJID[adae_saf$AESER == "Y" & adae_saf$TRT01P == "Docetaxel 75mg/m2"]
)

log_action(log,
  action = "compute_sae_summary",
  object = "SAE summary table",
  reason = sprintf(
    "Summarised per SAP Section 6.3. SAEs: %d subjects in KRASi arm (%.1f%%), %d in docetaxel arm (%.1f%%)",
    n_sae_krasi,
    n_sae_krasi / n_per_arm * 100,
    n_sae_doce,
    n_sae_doce / n_per_arm * 100
  )
)

saes |>
  select(Treatment = TRT01P, SOC = AEBODSYS, Term = AEDECOD, `N subjects` = n) |>
  head(10L) |>
  knitr::kable(caption = "Serious TEAEs (AESER = Y) — top 10")
```

---

## Pneumonitis signal review

Pneumonitis is a clinically important immune-mediated AE associated with KRAS inhibitors. We identify cases and document the medical review decision — three separate, individually logged steps that together form a single audit narrative.

```{r pneumonitis}
pneumonitis <- adae_saf |>
  filter(grepl("Pneumonitis", AEDECOD, ignore.case = TRUE))

n_pneu_krasi <- nrow(pneumonitis[pneumonitis$TRT01P == "KRASi 400mg QD", ])
n_pneu_doce  <- nrow(pneumonitis[pneumonitis$TRT01P == "Docetaxel 75mg/m2", ])
n_pneu_g3    <- nrow(pneumonitis[pneumonitis$AETOXGR %in% c("3","4"), ])

log_note(log,
  sprintf(
    "Safety signal identified: pneumonitis — %d cases in KRASi arm vs %d in
     docetaxel arm. Grade 3/4: %d cases. Flagged for medical monitor review
     per Safety Review Plan.",
    n_pneu_krasi, n_pneu_doce, n_pneu_g3
  )
)

# Log individual Grade 3+ pneumonitis cases
g3_pneu <- pneumonitis |> filter(AETOXGR %in% c("3","4"))
for (i in seq_len(nrow(g3_pneu))) {
  log_note(log,
    sprintf(
      "Pneumonitis case — USUBJID %s | Grade %s | Treatment: %s | Action: %s | Related: %s",
      g3_pneu$USUBJID[i], g3_pneu$AETOXGR[i],
      g3_pneu$TRT01P[i], g3_pneu$AEACN[i], g3_pneu$AEREL[i]
    )
  )
}

# Log medical monitor decision
log_note(log,
  sprintf(
    "Medical monitor reviewed all %d pneumonitis cases (ref:
     MM-review-20260620-001). Conclusion: incidence consistent with class
     effect; benefit-risk remains favourable. Risk mitigation: enhanced
     monitoring protocol added to IB (v4.2, dated 2026-06-22). No protocol
     amendment required at this time.",
    n_pneu_krasi
  )
)

pneumonitis |>
  select(USUBJID, Treatment = TRT01P, Term = AEDECOD,
         Grade = AETOXGR, Serious = AESER,
         Related = AEREL, Action = AEACN) |>
  knitr::kable(caption = "Pneumonitis cases — all grades")
```

---

## Discontinuations due to AEs

```{r discontinuations}
disc_ae <- adae_saf |>
  filter(AEACN == "DRUG WITHDRAWN") |>
  group_by(TRT01P, AEBODSYS, AEDECOD) |>
  summarise(n = n_distinct(USUBJID), .groups = "drop") |>
  arrange(desc(n))

n_disc_krasi <- n_distinct(
  adae_saf$USUBJID[adae_saf$AEACN == "DRUG WITHDRAWN" &
                   adae_saf$TRT01P == "KRASi 400mg QD"]
)
n_disc_doce <- n_distinct(
  adae_saf$USUBJID[adae_saf$AEACN == "DRUG WITHDRAWN" &
                   adae_saf$TRT01P == "Docetaxel 75mg/m2"]
)

log_action(log,
  action = "compute_disc_due_to_ae",
  object = "Discontinuation due to AE table",
  reason = sprintf(
    "Tabulated per SAP Section 6.4. Discontinued due to AE: %d subjects in KRASi arm (%.1f%%), %d in docetaxel (%.1f%%)",
    n_disc_krasi, n_disc_krasi / n_per_arm * 100,
    n_disc_doce,  n_disc_doce  / n_per_arm * 100
  )
)

disc_ae |>
  select(Treatment = TRT01P, SOC = AEBODSYS,
         Term = AEDECOD, `N subjects` = n) |>
  head(10L) |>
  knitr::kable(caption = "Discontinuations due to AE — top 10")
```

---

## Safety visualisations

```{r forest-plot}
teae_wide <- teae_summary |>
  filter(AEBODSYS %in% top_socs) |>
  select(TRT01P, AEBODSYS, pct) |>
  pivot_wider(names_from = TRT01P, values_from = pct, values_fill = 0) |>
  rename(KRASi = `KRASi 400mg QD`, Docetaxel = `Docetaxel 75mg/m2`)

teae_wide |>
  tidyr::pivot_longer(cols = c(KRASi, Docetaxel),
                      names_to = "Treatment", values_to = "pct") |>
  ggplot(aes(x = pct, y = reorder(AEBODSYS, pct), fill = Treatment)) +
  geom_col(position = "dodge", width = 0.65, alpha = 0.85) +
  scale_fill_manual(values = c("KRASi" = "#1a56db", "Docetaxel" = "#6b6f80")) +
  labs(
    title    = "TEAE Incidence by System Organ Class",
    subtitle = "Safety analysis set | Top 5 SOCs",
    x        = "Subjects with ≥1 TEAE (%)",
    y        = NULL,
    fill     = "Treatment"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")

log_note(log, "Figure generated: TEAE incidence by SOC (top 5), dodge bar chart")
```

```{r grade-plot}
adae_saf |>
  count(TRT01P, AETOXGR) |>
  group_by(TRT01P) |>
  mutate(pct = n / sum(n) * 100) |>
  ggplot(aes(x = AETOXGR, y = pct, fill = TRT01P)) +
  geom_col(position = "dodge", width = 0.65, alpha = 0.85) +
  scale_fill_manual(values = c("KRASi 400mg QD" = "#1a56db",
                               "Docetaxel 75mg/m2" = "#6b6f80")) +
  labs(
    title    = "AE Grade Distribution by Treatment Arm",
    subtitle = "Safety analysis set | % of all AEs",
    x        = "Grade (CTCAE)",
    y        = "% of AEs",
    fill     = "Treatment"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")

log_note(log, "Figure generated: AE grade distribution by treatment arm")
```

---

## Safety summary for SMC

```{r smc-summary}
smc_table <- data.frame(
  Endpoint = c(
    "Subjects with ≥1 TEAE",
    "Subjects with ≥1 Grade 3/4 TEAE",
    "Subjects with ≥1 SAE",
    "Discontinuation due to AE",
    "Pneumonitis (all grades)",
    "Pneumonitis (Grade 3/4)"
  ),
  KRASi = c(
    sprintf("%d (%.1f%%)", n_distinct(adae_saf$USUBJID[adae_saf$TRT01P=="KRASi 400mg QD"]),
            n_distinct(adae_saf$USUBJID[adae_saf$TRT01P=="KRASi 400mg QD"]) / n_per_arm * 100),
    sprintf("%d (%.1f%%)", n_distinct(adae_saf$USUBJID[adae_saf$TRT01P=="KRASi 400mg QD" & adae_saf$AETOXGR %in% c("3","4")]),
            n_distinct(adae_saf$USUBJID[adae_saf$TRT01P=="KRASi 400mg QD" & adae_saf$AETOXGR %in% c("3","4")]) / n_per_arm * 100),
    sprintf("%d (%.1f%%)", n_sae_krasi,  n_sae_krasi  / n_per_arm * 100),
    sprintf("%d (%.1f%%)", n_disc_krasi, n_disc_krasi / n_per_arm * 100),
    sprintf("%d (%.1f%%)", n_pneu_krasi, n_pneu_krasi / n_per_arm * 100),
    sprintf("%d (%.1f%%)", n_pneu_g3,    n_pneu_g3    / n_per_arm * 100)
  ),
  Docetaxel = c(
    sprintf("%d (%.1f%%)", n_distinct(adae_saf$USUBJID[adae_saf$TRT01P=="Docetaxel 75mg/m2"]),
            n_distinct(adae_saf$USUBJID[adae_saf$TRT01P=="Docetaxel 75mg/m2"]) / n_per_arm * 100),
    sprintf("%d (%.1f%%)", n_distinct(adae_saf$USUBJID[adae_saf$TRT01P=="Docetaxel 75mg/m2" & adae_saf$AETOXGR %in% c("3","4")]),
            n_distinct(adae_saf$USUBJID[adae_saf$TRT01P=="Docetaxel 75mg/m2" & adae_saf$AETOXGR %in% c("3","4")]) / n_per_arm * 100),
    sprintf("%d (%.1f%%)", n_sae_doce,  n_sae_doce  / n_per_arm * 100),
    sprintf("%d (%.1f%%)", n_disc_doce, n_disc_doce / n_per_arm * 100),
    sprintf("%d (%.1f%%)", n_pneu_doce, n_pneu_doce / n_per_arm * 100),
    "0 (0.0%)"
  ),
  stringsAsFactors = FALSE
)

knitr::kable(smc_table,
  col.names = c("Safety endpoint",
                "KRASi 400mg QD (N=200)",
                "Docetaxel 75mg/m2 (N=200)"),
  caption   = "Safety summary — NSCLC-KRASi-301 (data cut 2026-05-15)"
)

log_action(log,
  action = "smc_safety_table_compiled",
  object = "Integrated safety summary",
  reason = "Compiled for Safety Monitoring Committee. All 6 pre-specified safety endpoints computed and tabulated per SMC charter."
)
```

---

## Dual sign-off: biostatistician and medical monitor

`log_signature()` resolves the signer identity from the session `user` and cannot be overridden per call — by design, a signature always reflects who actually opened that session, not an arbitrary name passed at signing time.

Oncology safety reports typically require two independent sign-offs. Each call to `regulog_init()` writes a fresh genesis record and starts its own entry numbering, so two sessions cannot currently share a single `.rlog` file as one continuous chain. The correct pattern with the current package is two separate `.rlog` files — one per signer — each independently hash-chained and independently verifiable. Both reference the same analysis by `app` name and the same data cut, and both are bundled together for submission.

```{r signatures}
# Biostatistician sign-off — same session used throughout this analysis
log_signature(log,
  "I confirm that the safety analyses were conducted in accordance with
   SAP v2.1 and the SMC charter. All AEs, SAEs, and discontinuations are
   correctly classified and tabulated."
)

# Medical monitor sign-off — independent session, independent .rlog file
log_mm <- regulog_init(
  app     = "NSCLC-KRASi-301-safety-summary",
  version = "1.0.0",
  user    = "medical.monitor",
  path    = file.path(tempdir(), "audit_KRASi301_safety_mm_review_v1.rlog")
)

log_action(log_mm,
  action = "review_safety_summary",
  object = "NSCLC-KRASi-301 integrated safety summary",
  reason = "Reviewed all pneumonitis cases and the integrated safety
            summary produced in audit_KRASi301_safety_v1.rlog"
)

log_signature(log_mm,
  "I confirm that I have reviewed the safety data including all
   pneumonitis cases and the integrated safety summary. The
   benefit-risk profile remains favourable for KRASi 400mg QD."
)
```

---

## Verify and export

Each `.rlog` file is verified and exported independently. Both are bundled
together in the submission package.

```{r verify-export}
# Verify each session's chain independently
verify_log(log)
verify_log(log_mm)

# Export both for the submission package
trail <- export_audit_trail(log,
  format = "csv",
  signed = TRUE,
  path   = file.path(tempdir(), "audit_KRASi301_safety_v1.csv")
)

trail_mm <- export_audit_trail(log_mm,
  format = "csv",
  signed = TRUE,
  path   = file.path(tempdir(), "audit_KRASi301_safety_mm_review_v1.csv")
)

# Preview the primary analysis log
filter_log(log) |>
  head(12L) |>
  mutate(
    time   = substr(timestamp, 12L, 19L),   # HH:MM:SS only — full ISO timestamp is too wide
    action = ifelse(is.na(action), type, action)
  ) |>
  select(time, type, action, object) |>
  knitr::kable(caption = "Audit trail — first 12 entries")
```

---

## What this delivers to regulators

The exported audit trail documents the complete safety monitoring process:

- **Data provenance** — what datasets were loaded, from where, with row and column counts captured at the moment of read
- **Population decisions** — who was included in the safety set and why
- **Analysis execution** — every TEAE table, Grade 3/4 summary, and SAE listing was logged at computation
- **Safety signal handling** — the pneumonitis signal was identified, individually documented, reviewed by the medical monitor, and a decision reached — all in the analysis hash chain
- **Dual sign-off** — the biostatistician's analysis chain and the medical monitor's independent review chain are each separately tamper-evident and individually verifiable, bundled together as the complete sign-off record
- **Tamper evidence** — any post-hoc modification to any entry in either chain breaks its SHA-256 chain, detectable without reference to the original system

This directly satisfies 21 CFR Part 11 §11.10(b) (audit trails), §11.10(e) (tamper evidence), §11.100/§11.200 (electronic signature components), and EU Annex 11 Clause 9 (audit trail) — not through a parallel documentation process, but as an inherent property of the analysis itself.

---

## Session information

```{r session-info}
sessionInfo()
```
